## Chapter 2 problems

> 2E1. Which of the expressions below correspond to the statement:the probability of rain on Monday?

1. Pr(rain)
2. **Pr(rain|Monday)**
3. Pr(Monday|rain)
4. Pr(rain,Monday)/Pr(Monday)


> 2E2. Which of the following statements corresponds to the expression: Pr(Monday|rain)?

1. The probability of rain on Monday.
2. The probability of rain, given that it is Monday.
3. **The probability that it is Monday, given that it is raining.**
4. The probability that it is Monday and that it is raining.

> 2E3. Which of the expressions below correspond to the statement: the probability that it is Monday, given that it is raining?

1. **Pr(Monday|rain)**
2. Pr(rain|Monday)
3. Pr(rain|Monday)Pr(Monday)
4. Pr(rain|Monday)Pr(Monday)/Pr(rain) 
5. Pr(Monday|rain)Pr(rain)/Pr(Monday)

> 2E4. The Bayesian statistician Bruno de Finetti (1906–1985) began his book on probability theory with the declaration: “PROBABILITY DOES NOT EXIST.” The capitals appeared in the original, so I imagine de Finetti wanted us to shout this statement. What he meant is that probability is a device for describing uncertainty from the perspective of an observer with limited knowledge; it has no objective reality. Discuss the globe tossing example from the chapter, in light of this statement. What does it mean to say “the probability of water is 0.7”?

Based on the tosses that we've performed, on the globe we hold, we can expect 70% of future tosses to also land on water. The source of the uncertainty and limited knowledge is that the point we're landing on is random.


2M1. Recall the globe tossing model from the chapter. Compute and plot the grid approximate posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for p.

```{r}
globe_grid <- function(trial_list, grid_size){
  # define grid
  p_grid <- seq( from=0 , to=1 , length.out=grid_size ) 
  # define prior
  prior <- rep( 1 , grid_size )
  # compute likelihood at each value in grid 
  likelihood <- dbinom( sum(trial_list) , size=length(trial_list) , prob=p_grid )
  # compute product of likelihood and prior 
  unstd.posterior <- likelihood * prior
  # standardize the posterior, so it sums to 1 
  posterior <- unstd.posterior / sum(unstd.posterior)
  plot(p_grid , posterior , type="b" , 
       xlab="probability of water" , ylab="posterior probability")
  mtext( sprintf("%i points", grid_size ))  
}
```

1. W, W, W
```{r, fig.align='center'}
globe_grid(c(1,1,1),100)
```
2. W, W, W, L
```{r, fig.align='center'}
globe_grid(c(1,1,1,0),100)
```

3. L,W,W,L,W,W,W
```{r, fig.align='center'}
globe_grid(c(0,1,1,0,1,1,1),100)
```

2M2. Now assume a prior for p that is equal to zero when $p < 0.5$ and is a positive constant when $p \geq 0.5$. Again compute and plot the grid approximate posterior distribution for each of the sets of observations in the problem just above.

```{r, include=FALSE}
globe_grid_const <- function(trial_list, grid_size, prior_const){
  # define grid
  p_grid <- seq( from=0 , to=1 , length.out=grid_size ) 
  # define prior
  prior <- (p_grid >= 0.5) * prior_const
  # compute likelihood at each value in grid 
  likelihood <- dbinom( sum(trial_list) , size=length(trial_list) , prob=p_grid )
  # compute product of likelihood and prior 
  unstd.posterior <- likelihood * prior
  # standardize the posterior, so it sums to 1 
  posterior <- unstd.posterior / sum(unstd.posterior)
  plot(p_grid , posterior , type="b" , 
       xlab="probability of water" , ylab="posterior probability")
  mtext( sprintf("%i points", grid_size ))  
}
```
Change function prior definition to:
```
# define prior
prior <- (p_grid >= 0.5) * prior_const
```  
1. W, W, W
```{r, fig.align='center'}
globe_grid_const(c(1,1,1),100, 1.0)
```

2. W, W, W, L
```{r, fig.align='center'}
globe_grid_const(c(1,1,1,0),100, 1.0)
```


3. L,W,W,L,W,W,W
```{r, fig.align='center'}
globe_grid_const(c(0,1,1,0,1,1,1),100,1)
```



