## Chapter 4 problems

```{r, include=FALSE}
library(rethinking)
```

__4E1.__

>In the model definition below, which line is the likelihood?


\begin{align*}
y_i &\sim \text{Normal}(\mu,\sigma) \\
\mu &\sim \text{Normal}(0,10) \\
\sigma &\sim \text{Exponential}(1)
\end{align*}

Line 1, the $y_i$ definition, is the likelihood.

__4E2.__ 

>In the model definition just above, how many parameters are in the posterior distribution?

2 parameters ($\mu, \sigma$)

__4E3.__

> Using the model definition above, write down the appropriate form of Bayes’ theorem that includes the proper likelihood and priors.


TODO

__4E4.__

>In the model definition below, which line is the linear model?

\begin{align*}
  y_i &\sim \text{Normal}(\mu,\sigma)\\
  \mu_i &= \alpha + \beta x_i \\
  \alpha &\sim \text{Normal}(0,10)\\
  \beta &\sim \text{Normal}(0,1)\\
  \sigma &\sim \text{Exponential}(2)
\end{align*}

The second line, $\mu_i$ definition, is the linear model.

__4E5.__

> In the model definition just above, how many parameters are in the posterior distribution?

3 parameters ($\alpha, \beta, \sigma$)


__4M1.__

> For the model definition below, simulate observed y values from the prior (not the posterior).

\begin{align*}
  y_i &\sim \text{Normal}(\mu,\sigma)\\
  \mu &\sim \text{Normal}(0,10)\\
  \sigma &\sim \text{Exponential}(1)
\end{align*}


```{r, fig.height=3}
mu <- rnorm(10000, 0, 10) # normal from 0 to 10
sigma <- rexp(10000, 1) # Exponential rate 1
y_prior <- rnorm(10000, mu, sigma)
dens(y_prior)
```



__4M2.__

> Translate the model just above into a quap formula.

```{r}
flist <- alist(
  y ~ dnorm(mu, sigma),
  mu ~ dnorm(0,10),
  sigma ~ dexp(1)
  )
```

__4M3.__ 

> Translate the quap model formula below into a mathematical model definition.

```{r}
flist <- alist(
  y ~ dnorm( mu , sigma ), 
  mu <- a + b*x,
  a ~ dnorm( 0 , 10 ),
  b ~ dunif( 0 , 1 ), 
  sigma ~ dexp( 1 )
)
```

\begin{align*}
  y_i &\sim \text{Normal}(\mu,\sigma)\\
  \mu_i &= a + b x_i\\
  a &\sim \text{Normal}(0,10)\\
  b &\sim \text{Uniform}(0,1)\\
  \sigma &\sim \text{Exponential}(1)
\end{align*}


__4M4.__

> A sample of students is measured for height each year for 3 years. After the third year, you want to fit a linear regression predicting height using year as a predictor. Write down the mathematical model definition for this regression, using any variable names and priors you choose. Be prepared to defend your choice of priors.

\begin{align*}
  y_i &\sim \text{Normal}(\mu,\sigma)\\
  \mu_i &= \alpha + \beta x_i\\
  \alpha &\sim \text{Normal}(152,6)\\
  \beta &\sim \text{Uniform}(8,4)\\
  \sigma &\sim \text{Uniform}(0,30)
\end{align*}


Defense:

For simplicity, heights given as normal (typically might expect bimodal for male/female). Parameters are $\mu$ and $\sigma$. 

Use a linear regression to get $\mu$, where $x_i$ is year. $\sigma$, the deviation, we'll set to uniform. I'd expect about 95% (2$\sigma$) to fall within 1.3 ft (e.g. if mean is 5.5 ft, 95% will be between 6.15 ft and 4.85 ft), so that puts $\sigma$ at 0.65 ft $\sim$ 20 cm, so to be conservative, we'll do a uniform over 0-30 cm.

Set regression $\alpha$ to an average height; not indicating what age "students" are (could be elementary, could be college), so let's say 5 ft ($\sim 152$ cm), with a 0.65 ft ($\sim 20$ cm) deviation. Then $\beta$ is how much growth happens per year. Let's say a high schooler is full grown at 18, at 5.5 ft on average, born just under a foot, so over 18 years grow 4.5 foot, so that's about 3 inches a year, or just under $8$ cm, and give that a reasonable distribution of half that.


__4M5.__

> Now suppose I remind you that every student got taller each year. Does this information lead you to change your choice of priors? How?

The problem suggests that I might want to update $\beta$, but we've already considered that students are growing each year, so probably don't need to update the prior. We want to ensure that $\beta$ is never negative, so if anything we might reduce the standard deviation to remove the possibility of negative values there at the fringe of the tails.



__4M6.__

> Now suppose I tell you that the variance among heights for students of the same age is never more than 64cm. How does this lead you to revise your priors?

Variance is $\sigma^2$, so that makes the standard deviation for $\beta$ at most $\sqrt{64}=8$. I set this value to 4 already, implying my prior is a stronger constraint than the new information given. This might lead me to loosen that constraint a little bit, for safety.






