## Chapter 4 problems

```{r, include=FALSE}
library(rethinking)
```

__4E1.__

>In the model definition below, which line is the likelihood?


\begin{align*}
y_i &\sim \text{Normal}(\mu,\sigma) \\
\mu &\sim \text{Normal}(0,10) \\
\sigma &\sim \text{Exponential}(1)
\end{align*}

Line 1, the $y_i$ definition, is the likelihood.

__4E2.__ 

>In the model definition just above, how many parameters are in the posterior distribution?

2 parameters ($\mu, \sigma$)

__4E3.__

> Using the model definition above, write down the appropriate form of Bayes’ theorem that includes the proper likelihood and priors.


TODO

__4E4.__

>In the model definition below, which line is the linear model?

\begin{align*}
  y_i &\sim \text{Normal}(\mu,\sigma)\\
  \mu_i &= \alpha + \beta x_i \\
  \alpha &\sim \text{Normal}(0,10)\\
  \beta &\sim \text{Normal}(0,1)\\
  \sigma &\sim \text{Exponential}(2)
\end{align*}

The second line, $\mu_i$ definition, is the linear model.

__4E5.__

> In the model definition just above, how many parameters are in the posterior distribution?

3 parameters ($\alpha, \beta, \sigma$)


__4M1.__

> For the model definition below, simulate observed y values from the prior (not the posterior).

\begin{align*}
  y_i &\sim \text{Normal}(\mu,\sigma)\\
  \mu &\sim \text{Normal}(0,10)\\
  \sigma &\sim \text{Exponential}(1)
\end{align*}


```{r, fig.height=3}
mu <- rnorm(10000, 0, 10) # normal from 0 to 10
sigma <- rexp(10000, 1) # Exponential rate 1
y_prior <- rnorm(10000, mu, sigma)
dens(y_prior)
```



__4M2.__

> Translate the model just above into a quap formula.

```{r}
flist <- alist(
  y ~ dnorm(mu, sigma),
  mu ~ dnorm(0,10),
  sigma ~ dexp(1)
  )
```

__4M3.__ 

> Translate the quap model formula below into a mathematical model definition.

```{r}
flist <- alist(
  y ~ dnorm( mu , sigma ), 
  mu <- a + b*x,
  a ~ dnorm( 0 , 10 ),
  b ~ dunif( 0 , 1 ), 
  sigma ~ dexp( 1 )
)
```

\begin{align*}
  y_i &\sim \text{Normal}(\mu,\sigma)\\
  \mu_i &= a + b x_i\\
  a &\sim \text{Normal}(0,10)\\
  b &\sim \text{Uniform}(0,1)\\
  \sigma &\sim \text{Exponential}(1)
\end{align*}


__4M4.__

> A sample of students is measured for height each year for 3 years. After the third year, you want to fit a linear regression predicting height using year as a predictor. Write down the mathematical model definition for this regression, using any variable names and priors you choose. Be prepared to defend your choice of priors.

\begin{align*}
  y_i &\sim \text{Normal}(\mu,\sigma)\\
  \mu_i &= \alpha + \beta x_i\\
  \alpha &\sim \text{Normal}(152,6)\\
  \beta &\sim \text{Uniform}(8,4)\\
  \sigma &\sim \text{Uniform}(0,30)
\end{align*}


Defense:

For simplicity, heights given as normal (typically might expect bimodal for male/female). Parameters are $\mu$ and $\sigma$. 

Use a linear regression to get $\mu$, where $x_i$ is year. $\sigma$, the deviation, we'll set to uniform. I'd expect about 95% (2$\sigma$) to fall within 1.3 ft (e.g. if mean is 5.5 ft, 95% will be between 6.15 ft and 4.85 ft), so that puts $\sigma$ at 0.65 ft $\sim$ 20 cm, so to be conservative, we'll do a uniform over 0-30 cm.

Set regression $\alpha$ to an average height; not indicating what age "students" are (could be elementary, could be college), so let's say 5 ft ($\sim 152$ cm), with a 0.65 ft ($\sim 20$ cm) deviation. Then $\beta$ is how much growth happens per year. Let's say a high schooler is full grown at 18, at 5.5 ft on average, born just under a foot, so over 18 years grow 4.5 foot, so that's about 3 inches a year, or just under $8$ cm, and give that a reasonable distribution of half that.


__4M5.__

> Now suppose I remind you that every student got taller each year. Does this information lead you to change your choice of priors? How?

The problem suggests that I might want to update $\beta$, but we've already considered that students are growing each year, so probably don't need to update the prior. We want to ensure that $\beta$ is never negative, so if anything we might reduce the standard deviation to remove the possibility of negative values there at the fringe of the tails.



__4M6.__

> Now suppose I tell you that the variance among heights for students of the same age is never more than 64cm. How does this lead you to revise your priors?

Variance is $\sigma^2$, so that makes the standard deviation for $\beta$ at most $\sqrt{64}=8$. I set this value to 4 already, implying my prior is a stronger constraint than the new information given. This might lead me to loosen that constraint a little bit, for safety.

__4H1.__

>The weights listed below were recorded in the !Kung census, but heights were not recorded for these individuals. Provide predicted heights and 89% intervals for each of these individuals. That is, fill in the table below, using model-based predictions.

```{r}
# Load data
data("Howell1")
d <- Howell1

height_model <- map(
  alist(
    height ~ dnorm(a + b*weight, sigma),
    a ~ dnorm(178, 20),
    b ~ dnorm(0, 10),
    sigma ~ dunif(0, 50) # most pirors stolen from chapter
  ), 
  data = d
)
to_evaluate <- c(46.95, 43.72, 64.78, 32.59, 54.63)
simulated_heights <- sim(height_model, data=list(weight=to_evaluate), n=1e5, silent=TRUE)
mean_val <- apply(simulated_heights, # on heights
                  2, # Columns (not rows)
                  mean # Evaluate mean
                  )
print(mean_val)
```
```{r}
percentile_interval <- apply(simulated_heights,
                             2,
                             PI)
print(percentile_interval)
```


__4H2.__

> Select out all the rows in the Howell1 data with ages below 18 years of age. If you do it right, you should end up with a new data frame with 192 rows in it.

```{r}
d_youth <- d[d$age < 18,]
nrow(d_youth)
```

>(a) Fit a linear regression to these data, using quap. Present and interpret the estimates. For every 10 units of increase in weight, how much taller does the model predict a child gets?

```{r}

youth_height_model <- quap(
  alist(
    height ~ dnorm(a + b*weight, sigma),
    a ~ dnorm(178, 100),
    b ~ dnorm(0, 10),
    sigma ~ dunif(0, 50)
  ),
  data = d_youth
)

print(youth_height_model)
```

For every 10 units of increase in weight, height increases by about 27 cm.

> (b) Plot the raw data, with height on the vertical axis and weight on the horizontal axis. Superimpose the MAP regression line and 89% interval for the mean. Also superimpose the 89% interval for predicted heights.

```{r, warning=FALSE}
plot( height ~ weight , data=d_youth , col=col.alpha("firebrick",0.5) )
x_seq <- seq(from=min(d_youth$weight), to=max(d_youth$weight), by=0.5)

# Sample posterior

# Inital try, didn't work
# mu <- link(youth_height_model, data=data.frame(weight=x_seq))  

posterior <- extract.samples(youth_height_model)
mu_link_lambda <- function(weight){
  posterior$a + posterior$b*weight
} 
mu <- sapply(x_seq, mu_link_lambda)

mu.mean <- apply(mu, 2, mean )
mu.PI <- apply(mu, 2, PI)
mu.HPDI <- apply(mu, 2, HPDI)

simulated_heights <- sim(youth_height_model,data=list(weight=x_seq))

height.PI <- apply(simulated_heights, 2, PI)

# plot the MAP line, aka the mean mu for each weight 
lines( x_seq , mu.mean )
# plot a shaded region for 89% PI 
shade(mu.HPDI,x_seq, col=col.alpha('cornflowerblue',0.4))
shade( height.PI , x_seq,col=col.alpha('cornflowerblue',0.2))

```


>(c) What aspects of the model fit concern you? Describe the kinds of assumptions you would change, if any, to improve the model. You don’t have to write any new code. Just explain what the model appears to be doing a bad job of, and what you hypothesize would be a better model.

This data doesn't appear to be linear at the tails. Probably what the best method would be is to square the height values then draw a linear model on that:

```{r, warning=FALSE}
plot( height**2 ~ weight , data=d_youth , col=col.alpha("firebrick",0.5) )
```

Certainly looks more linear than before. Alternatively, a polynomial function would work too.






