## Chapter 9 problems

```{r, include=FALSE}
library(rethinking)
```

__9E1.__

>Which of the following is a requirement of the simple Metropolis algorithm?

1. The parameters must be discrete
2. The likelihood must be Gaussian
3. The proposal distributino must be symmetric

Only 3. Metropolis-Hastings generalizes for asymmetric distributions.

__9E2.__

>Gibbs sampling is more efficient than the Metropolis algorithm. How does it achieve this extra
efficiency? Are there any limitations to the Gibbs sampling strategy?

Gibbs uses conjugate pairs, which have analytical solutions, in order to have adaptive proposals for parameter values

__9E3.__ 

>Which sort of parameters can Hamiltonian Monte Carlo not handle? Can you explain why?

Discrete parameters. Since it simulates particles traversing a plane, it cannot deal with discrete parameters.

__9E4.__

> Explain the difference between the effective number of samples, n_eff as calculated by Stan, and the actual number of samples.

n_eff is a crude number of independent samples you get, which is fewer than the number thrown due to autocorrelation.

__9E5.__

>Which value should Rhat approach, when a chain is sampling the posterior distribution correctly?'

Should approach 1 from above 1.

__9E6.__

>Sketch a good trace plot for a Markov chain, one that is effectively sampling from the posterior distribution. What is good about its shape? Then sketch a trace plot for a malfunctioning Markov chain. What about its shape indicates malfunction?

A good one will be bounded and bounce around in a limited parameter space - "fuzzy caterpiller" I've seen described elsewhere. bad ones will jump to weird bad values.

TODO: Medium

__9H1.__

>Run the model below and then inspect the posterior distribution and explain what it is accomplishing.

```{r}
mp <- map2stan( 
  alist(
    a ~ dnorm(0,1),
    b ~ dcauchy(0,1) ),
    data=list(y=1),
    start=list(a=0,b=0),
    iter=1e4, warmup=100 , WAIC=FALSE )
```

```{r}
precis(mp)
```

```{r}
samples <- extract.samples(mp)
hist(samples$a)
```

```{r}
hist(samples$b)
```

$a$ is nice and symmetric. $b$ has a huge standard deviation, it's got several large values.

