# Chapter 8 - Conditional Manatees
```{r, include=FALSE}
library(rethinking)
```

Title explanation - manatees show damage from propeller blades they come in contact with, to avoid manatee deaths, you might consider propeller blade guards. Autopsies though show it's collisions with blunt sides of the boats that cause the deaths. Damage seen on living specimens _is conditional on survival_.

_Conditioning_ - is knowing how something occurred. Posteriors are conditional on data. Model-based inference is conditional on the model. Linear models, for example, allow each outcome to be conditional on a set of predictors.

_Interaction_ (also _moderation_) - a type of conditioning that allows parameters to be conditional on further aspects of data. Key to models beyond Gaussian outcomes and linear models.

Models with complex interactions are easy to fit to data, but hard to understand.

## 8.1 - Building an interaction

Example of Africa's log GDP vs terrain ruggedness - other nations have a negative relationship, Africa's is positive. One hypothesis is that rugged regions are protected against slave trades.

```{r, fig.height=2}
africa_dag <- dagitty( "dag {
  U[unobserved]
  R -> G <- C
  R <- U -> G
}")
coordinates( africa_dag ) <- list( x=c(R=0, G=1, U=1, C=2) ,
                                  y=c(R=0, G=0, U=1, C=0) )
drawdag( africa_dag )
```

$R$ is ruggedness, $G$ is GDP, $C$ is continent, $U$ is unobserved confounds (e.g. distance to coasts). $R$ and $C$ both influence $G$, could be independent or could interact; DAG can only imply to us that $G=f(R,C)$.

Want to build a model for $f(R,C)$. Could try splitting the dataframe into two, one for Africa and one not, but this is bad: some variables don't depend at all on continent, you can't get probability on the variable you used to split, you need all data to use information criteria, and there's advantages to borrowing information across categories.

### Making a Rugged Model

```{r}
library(rethinking) 
data(rugged)
d <- rugged
# make log version of outcome 
d$log_gdp <- log( d$rgdppc_2000 )
# extract countries with GDP data
dd <- d[ complete.cases(d$rgdppc_2000) , ]
# rescale variables
dd$log_gdp_std <- dd$log_gdp / mean(dd$log_gdp) 
dd$rugged_std <- dd$rugged / max(dd$rugged)
```

Usually - subtract mean and divided by standard deviations to make variable into z-score. Not done here

- Ruggedness divided by maximum to make it a percentage, 0 is flat, 1 is the most rugged

- log GDP is divided by the average, so a proportion of the international average: 1 is average, .8 is 80% of average

\begin{align*}
  log(y_i) &\sim \text{Normal}(\mu_i,\sigma)\\
  \mu_i &= \alpha + \beta(r_i - \bar{r})
\end{align*}

Here \bar{r} is the average ruggedness, 0.215 - most nations aren't so rugged.

Priors:

Intercept should be close to 1, since we scaled the mean to 1. Slope we can start with no bias, 0 and a deviation of 1. Sigma, we can do a broad exponential of 1.

\begin{align*}
  \alpha &\sim \text{Normal}(1,1)\\
  \beta &\sim \text{Normal}(0,1)\\
  \sigma &\sim \text{Exponential}(1)
\end{align*}

```{r}
m8.1 <- quap( 
  alist(
    log_gdp_std ~ dnorm( mu , sigma ) , 
    mu <- a + b*( rugged_std - 0.215 ) , 
    a ~ dnorm( 1 , 1 ) ,
    b ~ dnorm( 0 , 1 ) ,
    sigma ~ dexp( 1 ) 
  ) ,data=dd )
```

Looking at priors:

```{r}
set.seed(7)
prior <- extract.prior( m8.1 )
# set up the plot dimensions
plot( NULL , xlim=c(0,1) , ylim=c(0.5,1.5) ,
      xlab="ruggedness" , ylab="log GDP" ) 
abline( h=min(dd$log_gdp_std) , lty=2 ) 
abline( h=max(dd$log_gdp_std) , lty=2 )
# draw 50 lines from the prior
rugged_seq <- seq( from=-0.1 , to=1.1 , length.out=30 )
mu <- link( m8.1 , post=prior , data=data.frame(rugged_std=rugged_seq) ) 
for ( i in 1:50 ) lines( rugged_seq , mu[i,] , col=col.alpha("black",0.3) )
```


Pretty big realm of possibilities - dashed lines are maxes and mins, yet many lie outside: tighter $\alpha$ (remember, 95% in 2$\sigma$). Slopes also too variable - current have more than half with slope more than 0.6. Try instead $\alpha \sim \text{Normal}(0,0.1)$, and $\beta \sim \text{Normal}(0,0.3)$.

```{r}
m8.1 <- quap( 
  alist(
    log_gdp_std ~ dnorm( mu , sigma ) , 
    mu <- a + b*( rugged_std - 0.215 ) , 
    a ~ dnorm( 1 , 0.1 ) ,
    b ~ dnorm( 0 , 0.3 ) ,
    sigma ~ dexp(1) 
  ) , 
  data=dd 
)
precis( m8.1)
```

### Adding an indicator variable isn't enough

To build a model that allows for models in and outside to have different intercepts, must modify such that mean is conditional on continent:

$$
  \mu_i = \alpha_{\text{CID}[i]} + \beta(r_i - \bar{r})
$$

For CID being a continent ID.

```{r}
# make variable to index Africa (1) or not (2) 
dd$cid <- ifelse( dd$cont_africa==1 , 1 , 2 )
```

For categorical variable, add brackets:

```{r}
m8.2 <- quap( 
  alist(
    log_gdp_std ~ dnorm( mu , sigma ) ,
    mu <- a[cid] + b*( rugged_std - 0.215 ) , 
    a[cid] ~ dnorm( 1 , 0.1 ) ,
    b ~ dnorm( 0 , 0.3 ) ,
    sigma ~ dexp( 1 )
) , data=dd )
compare( m8.1 , m8.2 )
```

Continent variable seems to be picking up on useful info (standard error of 15, difference of 64).

```{r}
precis(m8.2, depth=2)
```


Slope a[1] is intercept for African nations, lower than a[2]

```{r}
post <- extract.samples(m8.2) 
diff_a1_a2 <- post$a[,1] - post$a[,2] 
PI( diff_a1_a2 )
```

Reliably below zero. Sampling posterior and computing means and CIs for different splits:

```{r}
rugged.seq <- seq( from=-0.1 , to=1.1 , length.out=30 )
# compute mu over samples, fixing cid=2 
mu.NotAfrica <- link( m8.2 ,
                      data=data.frame( cid=2 , rugged_std=rugged.seq ) )
# compute mu over samples, fixing cid=1 
mu.Africa <- link( m8.2 ,
                   data=data.frame( cid=1 , rugged_std=rugged.seq ) )
# summarize to means and intervals
mu.NotAfrica_mu <- apply( mu.NotAfrica , 2 , mean )
mu.NotAfrica_ci <- apply( mu.NotAfrica , 2 , PI , prob=0.97 ) 
mu.Africa_mu <- apply( mu.Africa , 2 , mean )
mu.Africa_ci <- apply( mu.Africa , 2 , PI , prob=0.97 )
```

Drawing these (not done here) shows a weak negative relationship between economic development and ruggedness.

**All including a dummy variable has done is allow the model to predict a lower mean for the African nations, it did nothing for slope**. This is a bad model.

### Adding an interaction does work

Make slope conditional on continent:

$$
\mu_i = \alpha_{\text{CID}[i]} + \beta_{\text{CID}[i]}(r_i-\bar{r})
$$

Where $A_i$ is 0/1 for African nations.

```{r}
  
m8.3 <- quap( 
  alist(
    log_gdp_std ~ dnorm( mu , sigma ) ,
    mu <- a[cid] + b[cid]*( rugged_std - 0.215 ) , 
    a[cid] ~ dnorm( 1 , 0.1 ) ,
    b[cid] ~ dnorm( 0 , 0.3 ) ,
    sigma ~ dexp( 1 )
) , data=dd )
precis(m8.3, depth=2)
```

```{r}
compare( m8.1 , m8.2 , m8.3 , func=PSIS )
```

High weight for m8.3, but weight given to m8.2 suggests there's a chance slopes i m8.3 are overfit.

```{r}
plot( PSIS( m8.3 , pointwise=TRUE )$k )
```

Some values here are overweighted.


### Plotting the interaction

```{r}
d.A1 <- dd[ dd$cid==1 , ]
plot( d.A1$rugged_std , d.A1$log_gdp_std , pch=16 , col=rangi2 ,
    xlab="ruggedness (standardized)" , ylab="log GDP (as proportion of mean)" ,
    xlim=c(0,1) )
mu <- link( m8.3 , data=data.frame( cid=1 , rugged_std=rugged_seq ) )
mu_mean <- apply( mu , 2 , mean )
mu_ci <- apply( mu , 2 , PI , prob=0.97 )
lines( rugged_seq , mu_mean , lwd=2 )
shade( mu_ci , rugged_seq , col=col.alpha(rangi2,0.3) )
mtext("African nations")

# plot non-Africa - cid=2
d.A0 <- dd[ dd$cid==2 , ]
plot( d.A0$rugged_std , d.A0$log_gdp_std , pch=1 , col="black" ,
    xlab="ruggedness (standardized)" , ylab="log GDP (as proportion of mean)" ,
    xlim=c(0,1) )
mu <- link( m8.3 , data=data.frame( cid=2 , rugged_std=rugged_seq ) )
mu_mean <- apply( mu , 2 , mean )
mu_ci <- apply( mu , 2 , PI , prob=0.97 )
lines( rugged_seq , mu_mean , lwd=2 )
shade( mu_ci , rugged_seq )
mtext("Non-African nations")
```

Now we see the effect - slope reversal in Africa.


## 8.2 - Symmetry of Interactions

Interaction has two equal phrasings:

1. How much does association between ruggedness and log GDP depend on whether the nation is Africa?
2. How much does the association of Africa with log GDP depend on ruggedness?

Different to us, same to model.


```{r}
rugged_seq <- seq(from=-0.2,to=1.2,length.out=30)
muA <- link( m8.3 , data=data.frame(cid=1,rugged_std=rugged_seq) ) 
muN <- link( m8.3 , data=data.frame(cid=2,rugged_std=rugged_seq) ) 
delta <- muA - muN
```

This delta is counterfactual, looking at how the model compares nations inside and outside ruggedness - above .8, a nation would have better GDP if it were in Africa than not.

Simple interactions are symmetric, and within model context, no reason to prefer one interpretation over another.

## 8.3 - Continuous Interactions

Interaction effects are difficult to interpret, since multiple parameters are in play at a time.

Looking at a two-way interaction between continuous variables

### A winter flower

Looking at size of tulip bulbs under different light conditions

```{r}
library(rethinking) 
data(tulips)
d <- tulips
str(d)
```

Predicting blooms. "Water" and "shade" are low (1) to high (3).

### The Models

Looking at 2 - water and shade with and without interacting. Causally $W \rightarrow B \leftarrow S$, but doesn't tell us the function in which they influence $B$, $B = f(W,S)$.

Main effects model:

\begin{align*}
  B_i &\sim  \text{Normal}(\mu_i, \sigma)\\
  \mu_i &= \alpha \beta_W(W_i - \bar{W}) + \beta_S(S_i-\bar{S})
\end{align*}

Just linear regression subtracting means. $W$ is water, $S$ is shade. Blooms we scale 0-1 since it has large values on the raw scale, but want to keep 0 as a meaningful boundary.

```{r}
d$blooms_std <- d$blooms / max(d$blooms) 
d$water_cent <- d$water - mean(d$water) 
d$shade_cent <- d$shade - mean(d$shade)
```

Looking at priors:

\begin{align*}
  \alpha &\sim  \text{Normal}(0.5,1)\\
  \beta_W &\sim \text{Normal}(0,1)\\
  \beta_S &\sim \text{Normal}(0,1)\\
\end{align*}

as a starting point. This neglects basic botany (positive w/water, negative w/shade). Standard deviations also too broad

```{r}
a <- rnorm( 1e4 , 0.5 , 1 )
sum( a < 0 | a > 1 ) / length( a )
```

Most of the probability is outside the range of the data. Consider a tighter 0.25 deviation

```{r}
a <- rnorm( 1e4 , 0.5 , 0.25 ) 
sum( a < 0 | a > 1 ) / length( a )
```

would be better.

```{r}
m8.4 <- quap( 
  alist(
    blooms_std ~ dnorm( mu , sigma ) ,
    mu <- a + bw*water_cent + bs*shade_cent , 
    a ~ dnorm( 0.5 , 0.25 ) ,
    bw ~ dnorm( 0 , 0.25 ) ,
    bs ~ dnorm( 0 , 0.25 ) ,
    sigma ~ dexp( 1 )
) , data=d )
```

**Next** the interaction model

\begin{align*}
  B_i &\sim  \text{Normal}(\mu_i, \sigma)\\
  \mu_i &= \alpha \beta_W W_i + \beta_S S_i + \beta_{WS} W_i S_i
\end{align*}

Last a prior for the interaction term is tough. Try setting $\beta_WS$ to have same std as $\beta_W$.

```{r}
m8.5 <- quap(
    alist(
        blooms_std ~ dnorm( mu , sigma ) ,
        mu <- a + bw*water_cent + bs*shade_cent + bws*water_cent*shade_cent ,
        a ~ dnorm( 0.5 , 0.25 ) ,
        bw ~ dnorm( 0 , 0.25 ) ,
        bs ~ dnorm( 0 , 0.25 ) ,
        bws ~ dnorm( 0 , 0.25 ) ,
        sigma ~ dexp( 1 )
    ) , data=d )
```

### Plotting posterior predictions

First plot the non-interaction model:

For interaction, we try a _triptych_ plot, 3 plots in a single panel designed to understand interaction impacts.

```{r}
par(mfrow=c(1,3)) # 3 plots in 1 row
for ( s in -1:1 ) {
    idx <- which( d$shade_cent==s )
    plot( d$water_cent[idx] , d$blooms_std[idx] , xlim=c(-1,1) , ylim=c(0,1) ,
        xlab="water" , ylab="blooms" , pch=16 , col=rangi2, main=paste("Shade = ",s) )
    mu <- link( m8.4 , data=data.frame( shade_cent=s , water_cent=-1:1 ) )
    for ( i in 1:20 ) lines( -1:1 , mu[i,] , col=col.alpha("black",0.3) )
}
```

Shown for each value of shade - note slopes don't change, since these don't have the interaction term.

```{r}
par(mfrow=c(1,3)) # 3 plots in 1 row
for ( s in -1:1 ) {
    idx <- which( d$shade_cent==s )
    plot( d$water_cent[idx] , d$blooms_std[idx] , xlim=c(-1,1) , ylim=c(0,1) ,
        xlab="water" , ylab="blooms" , pch=16 , col=rangi2, main=paste("Shade = ",s) )
    mu <- link( m8.5 , data=data.frame( shade_cent=s , water_cent=-1:1 ) )
    for ( i in 1:20 ) lines( -1:1 , mu[i,] , col=col.alpha("black",0.3) )
}
```

Note here, once adding the interaction, the slopes do change. As shade increases, water effect decreases. Likely explanation is that plants need both water and sunlight to grow; at higher light levels, water can matter more since tulips have enough light to produce blooms.

### Plotting prior predictions

```{r}
set.seed(7)
prior <- extract.prior(m8.4)
par(mfrow=c(1,3)) # 3 plots in 1 row
for ( s in -1:1 ) {
    idx <- which( d$shade_cent==s )
    plot( d$water_cent[idx] , d$blooms_std[idx] , xlim=c(-1,1) , ylim=c(0,1) ,
        xlab="water" , ylab="blooms" , pch=16 , col=rangi2, main=paste("Shade = ",s) )
    mu <- link( m8.4 , data=data.frame( shade_cent=s , water_cent=-1:1 ),post=prior )
    for ( i in 1:20 ) lines( -1:1 , mu[i,] , col=col.alpha("black",0.3) )
}

prior <- extract.prior(m8.5)
par(mfrow=c(1,3)) # 3 plots in 1 row
for ( s in -1:1 ) {
    idx <- which( d$shade_cent==s )
    plot( d$water_cent[idx] , d$blooms_std[idx] , xlim=c(-1,1) , ylim=c(0,1) ,
        xlab="water" , ylab="blooms" , pch=16 , col=rangi2, main=paste("Shade = ",s) )
    mu <- link( m8.5 , data=data.frame( shade_cent=s , water_cent=-1:1 ), post=prior )
    for ( i in 1:20 ) lines( -1:1 , mu[i,] , col=col.alpha("black",0.3) )
}
```

Difficult to necessarily interpret, but if you take the same model parameters from the non-interaction plot, you get no change in slope; if you do from the interaction plot, you do get a change in slope.

## 8.4 - Summary

We looked at interactions. You can't see them in DAGs, so they're a bit tricky, but they're important for accurate predictions. Interpretation is also tough, _triptych_ plots help with this.

